---
title: "Assignment 02"
output:
  pdf_document: default
  word_document: default
---
Answers start at the second page, this page only shows how I loaded the data.
```{r}
library(readxl)
library(tidyverse)
library(matrixStats)
survival <- read_excel("C:/Users/John/Desktop/STAT 445/Data/survival_data.xlsx", col_names = F)
colnames(survival)=c("Group","B","C","D","E","F")
```
\newpage
Question 01 \

(a) Boxplot of the data\
```{r}
boxplot(survival[,2:6])
```
Scale can be an issue as value ranges in B and F are much smaller than the rest of the group.
And in group C, D, and E there are values that are significantly larger than the rest of the group,
means those outliers can potentially affect the correlations. \
 \
(b) Bivariate scatter matrix plot \
```{r}
plot(survival[,2:6], col=survival$Group, pch = 16)
```
Legend: B = black, C = red , D = green, E = blue, F = cyan \
\
(c) Sample mean vector :

\begin{align}
  \tilde{\bar{x}} &= \begin{pmatrix}
           63.24528 \\
           252.50943 \\
           239.79245 \\
           165.43396 \\
           35.69811 \\
       \end{pmatrix}
\end{align}
\
Sample standard deviation vector:

\begin{align}
  \tilde{s} &= \begin{pmatrix}
           10.15629 \\
           195.90614 \\
           182.24427 \\
           139.15425 \\
           19.21396 \\
       \end{pmatrix}
\end{align}
\
(d)Matrix $\bar{X^*}$ of samples means using the standardized values 

\begin{align}
  \bar{X^*} &= \begin{bmatrix}
  -0.15841606 & -0.4861344 & -0.3385651 & -0.26901056 & -0.04106505 \\
  0.20969439 & -0.4154256 & -0.6484015 & -0.18951245 & -0.18271157 \\
  0.06024443 & 0.2478840 & 0.4941976 & 0.38237349 & -0.03261611 \\
  -0.22107316 & 0.6312065 & 0.4638877 & 0.16456107 & 0.21521954 \\
  -0.18825277 & 0.7894456 & 0.7327576 & -0.05821331 & 0.42340149 \\
  \end{bmatrix}
\end{align}
\
(e) Distance matrix $\bar{D^*}$

\begin{align}
  \bar{D^*} &= \begin{bmatrix}
  0.0000000 & 0.5127254 & 1.3055503 & 1.4662787 & 1.7423809 \\
  0.5127254 & 0.0000000 & 1.4551391 & 1.6738889 & 1.9754213 \\
  1.3055503 & 1.4551391 & 0.0000000 & 0.5795336 & 0.9022303 \\
  1.4662787 & 1.6738889 & 0.5795336 & 0.0000000 & 0.4374654 \\
  1.7423809 & 1.9754213 & 0.9022303 & 0.4374654 & 0.0000000 \\
  \end{bmatrix}
\end{align}
\
(f) From the distance matrix we can see that, there is a relatively low separation between group 1 and 2,
group 3 and 4, group 4 and 5. There is a relatively large separation between group 1 and 5, group 
2 and 4, group 2 and 5. \
\
\
Code used to solve (c), (d), (e)
```{r}
# c
sample_mean <- colMeans(survival[,2:6])
sample_mean
sample_sd <- colSds(as.matrix(survival[,2:6]))
sample_sd
# d
X_bar <- matrix(data = NA, nrow = 5,ncol = 5)
for (i in 1:5) {
  group_data <- filter(survival, Group == i)
  X_bar[i,] <- colMeans(group_data[,2:6])
}
X_star <- matrix(data=NA, nrow=5, ncol=5)
for (j in 1:5) { # j represents each variable of X1,..,X9
  X_star[,j] <- (X_bar[,j]-sample_mean[j])/sample_sd[j]
}
X_star

# e
dist(X_star, method="euclidean",diag=T,upper=T)
```

\newpage
Question 2 \
\
(a) $A = corr(B)$

\begin{align}
  A &= \begin{bmatrix}
  1.0000000 & 0.9410886 & 0.8707802 & 0.8091758 & 0.7815510 & 0.7278784 & 0.6689597 \\
  0.9410886 & 1.0000000 & 0.9088096 & 0.8198258 & 0.8013282 & 0.7318546 & 0.6799537 \\
  0.8707802 & 0.9088096 & 1.0000000 & 0.8057904 & 0.7197996 & 0.6737991 & 0.6769384 \\
  0.8091758 & 0.8198258 & 0.8057904 & 1.0000000 & 0.9050509 & 0.8665732 & 0.8539900 \\
  0.7815510 & 0.8013282 & 0.7197996 & 0.9050509 & 1.0000000 & 0.9733801 & 0.7905565 \\
  0.7278784 & 0.7318546 & 0.6737991 & 0.8665732 & 0.9733801 & 1.0000000 & 0.7987302 \\
  0.6689597 & 0.6799537 & 0.6769384 & 0.8539900 & 0.7905565 & 0.7987302 & 1.0000000 \\
  \end{bmatrix}
\end{align}

\
(b) $det(A) = 9.011147 \times 10^{-6}$ 
\
(c) Since $A$ is the correlation matrix of the data, as $det(A)$ is getting closer to 0, it shows there's a stronger relationship 
between the variables. When $A$ is an identity matrix, that is, all the correlations equals $0$($r_{ij} = 0$ for $i\not=j$), $det(A)$ has the maximum value at $1$.
\
(d)

\begin{align}
  A^{-1}\tilde{b} &= \begin{pmatrix}
           0.29123846 \\
           0.01544922 \\
           0.41095061 \\
           -0.31259312 \\
           -0.17544947 \\
           0.56607556 \\
           0.46999267 \\
       \end{pmatrix}
\end{align}

\
(e)

\begin{align}
  \tilde{y} &= \begin{pmatrix}
           144.4946 \\
           146.5773 \\
           140.7780 \\
           150.9731 \\
           148.8515 \\
           143.8101 \\
           135.8276 \\
       \end{pmatrix}
\end{align}
\
(f) Projection of $\tilde{x}$ onto $\tilde{y}$
\begin{align}
  \begin{pmatrix}
           0.1270371 \\
           0.1288682 \\
           0.1237696 \\
           0.1327329 \\
           0.1308677 \\
           0.1264354 \\
           0.1194173 \\
       \end{pmatrix}
\end{align}
\newpage
Code used to compute the answers for (a) to (f):

```{r}
B <- read_excel("C:/Users/John/Desktop/STAT 445/Data/w-nat-track-rec.xlsx", col_names = F)
# a
A <- cor(B)
A

# b
det(A)

# d
b = c(1,1,1,1,1,1,1)
solve(A)%*%b

# e
x = c(1,-1,1,-1,1,-1,1)
y <- A%*%A%*%A%*%A%*%x
y

# f
# sum(x*y) is the dot product of x and y, sum(y^2) is the square of the norm of y
projection <-(sum(x*y)/sum(y^2))* y 
projection

```
\newpage

Question 3 \
\
(a) Please note that the columns in matrix $C$ are ordered by their corresponding eigenvalues in descending order

\begin{align}
  C &= \begin{bmatrix}
  -0.3777657 & -0.4071756 & -0.1405803 &  0.58706293 & -0.16706891 &  0.53969730 &  0.08893934 \\
  -0.3832103 & -0.4136291 & -0.1007833 &  0.19407501 &  0.09350016 & -0.74493139 & -0.26565662 \\
  -0.3680361 & -0.4593531 &  0.2370255 & -0.64543118 &  0.32727328 &  0.24009405 &  0.12660435 \\
  -0.3947810 &  0.1612459 &  0.1475424 & -0.29520804 & -0.81905467 & -0.01650651 & -0.19521315 \\
  -0.3892610 &  0.3090877 & -0.4219855 & -0.06669044 &  0.02613100 & -0.18898771 &  0.73076817 \\
  -0.3760945 &  0.4231899 & -0.4060627 & -0.08015699 &  0.35169796 &  0.24049968 & -0.57150644 \\
  -0.3552031 &  0.3892153 &  0.7410610 &  0.32107640 &  0.24700821 & -0.04826992 &  0.08208401 \\
  \end{bmatrix}
\end{align}

\begin{align}
  D &= \begin{bmatrix}
  5.807624 & 0.0000000 & 0.0000000 & 0.0000000 & 0.00000000 & 0.00000000 & 0.00000000 \\
  0.000000 & 0.6286934 & 0.0000000 & 0.0000000 & 0.00000000 & 0.00000000 & 0.00000000 \\
  0.000000 & 0.0000000 & 0.2793346 & 0.0000000 & 0.00000000 & 0.00000000 & 0.00000000 \\
  0.000000 & 0.0000000 & 0.0000000 & 0.1245547 & 0.00000000 & 0.00000000 & 0.00000000 \\
  0.000000 & 0.0000000 & 0.0000000 & 0.0000000 & 0.09097174 & 0.00000000 & 0.00000000 \\
  0.000000 & 0.0000000 & 0.0000000 & 0.0000000 & 0.00000000 & 0.05451882 & 0.00000000 \\
  0.000000 & 0.0000000 & 0.0000000 & 0.0000000 & 0.00000000 & 0.00000000 & 0.01430226 \\
  \end{bmatrix}
\end{align}

\
(b)
\begin{align}
  CC^{T} &= \begin{bmatrix}
   1.000000 &  0.0000000 & -0.0000000 &  0.0000000 &  0.00000000 &  0.00000000 & -0.00000000 \\ 
   0.000000 &  1.0000000 & -0.0000000 &  0.0000000 &  0.00000000 &  0.00000000 &  0.00000000 \\
  -0.000000 & -0.0000000 &  1.0000000 &  0.0000000 &  0.00000000 &  0.00000000 &  0.00000000 \\
   0.000000 &  0.0000000 &  0.0000000 &  1.0000000 &  0.00000000 & -0.00000000 &  0.00000000 \\
   0.000000 &  0.0000000 &  0.0000000 &  0.0000000 &  1.00000000 & -0.00000000 & -0.00000000 \\
   0.000000 &  0.0000000 &  0.0000000 & -0.0000000 & -0.00000000 &  1.00000000 & -0.00000000 \\
  -0.000000 &  0.0000000 &  0.0000000 &  0.0000000 & -0.00000000 & -0.00000000 &  1.00000000 \\
  \end{bmatrix}
\end{align}

\
(c)

\begin{align}
  C_1 &= \begin{bmatrix}
    -0.3777657 & -0.4071756 & -0.1405803 &  0.58706293 \\
    -0.3832103 & -0.4136291 & -0.1007833 &  0.19407501 \\
    -0.3680361 & -0.4593531 &  0.2370255 & -0.64543118 \\
    -0.3947810 &  0.1612459 &  0.1475424 & -0.29520804 \\
    -0.3892610 &  0.3090877 & -0.4219855 & -0.06669044 \\
    -0.3760945 &  0.4231899 & -0.4060627 & -0.08015699 \\
    -0.3552031 &  0.3892153 &  0.7410610 &  0.32107640 \\
  \end{bmatrix}
\end{align}

\begin{align}
  D_1 &= \begin{bmatrix}
    5.807624 & 0.0000000 & 0.0000000 & 0.0000000 \\
    0.000000 & 0.6286934 & 0.0000000 & 0.0000000 \\
    0.000000 & 0.0000000 & 0.2793346 & 0.0000000 \\
    0.000000 & 0.0000000 & 0.0000000 & 0.1245547 \\
  \end{bmatrix}
\end{align}

\
\
(d) 
\begin{align}
  A - C_1DC_1^T = 
  \begin{bmatrix} 
     0.0185322 & -0.0236776 &  0.0022514 &  0.0117144 & -0.0050283 &  0.0010041 & -0.0050700 \\
    -0.0236776 &  0.0320584 & -0.0074482 & -0.0055547 &  0.0051210 & -0.0046044 &  0.0037495 \\
     0.0022514 & -0.0074482 &  0.0131158 & -0.0249549 & -0.0003726 &  0.0125842 &  0.0068709 \\
     0.0117144 & -0.0055547 & -0.0249549 &  0.0615883 & -0.0038173 & -0.0248261 & -0.0185905 \\
    -0.0050283 &  0.0051210 & -0.0003726 & -0.0038173 &  0.0096471 & -0.0076151 &  0.0019424 \\
     0.0010041 & -0.0046044 &  0.0125842 & -0.0248261 & -0.0076151 &  0.0190772 &  0.0065991 \\
    -0.0050700 &  0.0037495 &  0.0068709 & -0.0185905 &  0.0019424 &  0.0065991 &  0.0057739 \\
  \end{bmatrix}
\end{align}
$||A - C_1DC_1^T|| = 0.09097174$
\
\
Code used to compute the answers for (a) to (d):

```{r}
# a
spectral_decomp <- eigen(A,symmetric=T,only.values=F)
C <- spectral_decomp$vectors
C

eigenvalues <- spectral_decomp$values
D <- diag(eigenvalues) 
D

# b
round(C%*%t(C),6)

# c
C1 <- C[,1:4]
C1
D1 <- D[1:4,1:4]
D1

# d
A-C1%*%D1%*%t(C1)

norm(A-C1%*%D1%*%t(C1),type="2")
```









