---
title: "Assignment04 P01"
author: "Dawu Liu"
output:
  pdf_document: default
  word_document: default
---
```{r, echo=FALSE, results='hide'}
library(MESS)
X <- read.table("C:/Users/John/Desktop/STAT 445/Data/assignment4_data1.txt", sep = ",")
#a
S <- cov(X) ; x_bar <- colMeans(X)
S; x_bar
```
\
(a)
Sample covariance matrix:
\begin{align} 
S &= \begin{bmatrix}  0.65821889919388 & 0.54853765690655 & 0.175868141491043 \\  0.54853765690655 & 2.32605135623547 & 0.0219115575942205 \\  0.175868141491043 & 0.0219115575942205 & 3.06394221761802 \\  
\end{bmatrix} 
\end{align}
mean vector:
\begin{align} 
\tilde{\bar{x}} &= \begin{pmatrix}  
0.1441416 \\  0.8633031 \\ -0.1925616 
\end{pmatrix} 
\end{align}

\
(b)
Matrix scatter plot with histograms being the diagonals, the numbers shown on the plot are the row numbers (observations).

```{r, echo=FALSE, results='hide',fig.height=5, fig.width=6.67}
#b
par(mfcol=c(3, 3), mar=c(2,2,2,2))
for(i in 1:3){
  for(j in 1:3) {
    if(i != j){
      plot(X[[i]],X[[j]],type="n", xlab = paste0("X", i), ylab = paste0("X", j))
      text(X[[i]],X[[j]])}
    else{
      hist(X[[i]], xlab = paste0("X", i), main = paste0("Histogram of X", i), prob=TRUE,cex.main=0.8)
      lines(density(X[[i]]))}
  }
}
```

\newpage
Univariate q-q plots:

```{r, echo=FALSE, results='hide',fig.height=4, fig.width=12}
par(mfrow=c(1, 3))
q <- vector("list", length = 3)
for (i in 1:3) {
  q[[i]]=qqnorm(X[[i]],type="n", main = paste0("Normal Q-Q Plot of X",i));qqline(X[[i]],col="red")
  text(q[[i]]$x, q[[i]]$y,  pos = 1, cex=1.5)
}
```
\
(c) Statistical distance vector $\tilde{D}$ and it's values versus index plot:
$\tilde{D}$  = 
```{r, echo=FALSE, fig.height=5, fig.width=10}
#c
D <- c() 
for (i in 1:nrow(X)) {
  x_i <- t(X[i,])
  D[i]=t(x_i-x_bar)%*%solve(S)%*%(x_i-x_bar)
}
round(D,3)
plot(D, main = "D values vs index", type="n");text(D)
```
\
(d) 
The 1st, 2nd, and 3rd variables are represented by $X_1$, $X_2$, and $X_3$ respectively. \
The scatterplot indicates the outliers are observation 12 and 54 in $X_1$ vs $X_2$; 
5 in $X_1$ vs $X_3$; 5, 12 and 54 in $X_2$ vs $X_3$.
Even 59 appears to be a bit far away from the cloud, it still follows the linear trend.
By looking at the q-q plots, 12 and 54 are significantly above the straight line in $X_2$, 
so as 5 for $X_3$. The $\tilde{D}$ values vs index plot shows that observation 5, 12, and 54 have the top
three largest distances. Thus we conclude the outliers are observation $\textbf{5, 12, 54}$.


```{r, echo=FALSE, results='hide'}
#e
new_X <- X[-c(5,12,54),]
S_new = cov(new_X)
S_new; round(S-S_new,4)
new_x_bar = colMeans(new_X)
new_x_bar; round(x_bar-new_x_bar,4)
```
\
(e) New sample covariance matrix:
\begin{align} 
S &= \begin{bmatrix}  0.668136852725533 & 0.517193086605832 & 0.227261815320189 \\  0.517193086605832 & 0.937622838632426 & 0.263446058804402 \\  0.227261815320189 & 0.263446058804402 & 0.499434430158578 \\  
\end{bmatrix} 
\end{align}

New mean vector:
\begin{align} 
\tilde{\bar{x}} &= \begin{pmatrix}  
0.1453851 \\ 0.7103047 \\ -0.3447489
\end{pmatrix} 
\end{align}

Subtract the new covariance matrix from the old covariance matrix, we get:
\begin{align} 
\text{difference, old - new} &= \begin{bmatrix}  -0.0099 & 0.0313 & -0.0514 \\  0.0313 & 1.3884 & -0.2415 \\  -0.0514 & -0.2415 & 2.5645 \\  
\end{bmatrix} 
\end{align}

Removing the outliers affects the covariances across all three variables. The biggest covariance change is between the 2nd and 3rd variables.
It also significantly decreases the variances of the 2nd and 3rd variables themselves. 

Subtract the new mean vector from the old mean vector, we get:
\begin{align} 
\text{difference, old - new} &= \begin{pmatrix}  
-0.0012 \\ 0.1530 \\ 0.1522 
\end{pmatrix} 
\end{align}

There is a significant decrease in the mean of the 2nd and 3rd variables (by 0.15). 

\
(f) Matrix scatter plot and univariate q-q plots for the new data set:  
```{r, echo=FALSE, results='hide'}
#f
par(mfcol=c(3, 3), mar=c(2,2,2,2))
for(i in 1:3){
  for(j in 1:3) {
    if(i != j){
      plot(new_X[[i]],new_X[[j]],xlab = paste0("X", i), ylab = paste0("X", j))
    }
    else{
      hist(new_X[[i]], xlab = paste0("X", i), main = paste0("Histogram of X", i), prob=TRUE,cex.main=0.8)
      lines(density(X[[i]]))}
  }
}
```

```{r, echo=FALSE, results='hide',fig.height=4, fig.width=12}
par(mfrow=c(1, 3))
for (i in 1:3) {
  qqnorm(new_X[[i]], main = paste0("Normal Q-Q Plot of new X",i));qqline(X[[i]],col="red")
}

```
This evidence is fairly consistent with a normal distribution. The histograms of all three variables appear to be normal.
The data points in matrix scatter plots for each of all three variables appear to fit in an ellipse pretty well, suggesting each pair of the
the data is bivariate normal. Also, in the q-q plots, the points in all three variables appear to fit the straight lines well, only
the third variable being slightly light tailed.

\
Code used to solve the questions(graphs are hidden):
```{r, fig.show='hide'}
rm(list = ls())
library(MESS)
X <- read.table("C:/Users/John/Desktop/STAT 445/Data/assignment4_data1.txt", sep = ",")
#a
S <- cov(X) ; x_bar <- colMeans(X)
S; x_bar

#b
par(mfcol=c(3, 3), mar=c(2,2,2,2))
for(i in 1:3){
  for(j in 1:3) {
    if(i != j){
      plot(X[[i]],X[[j]],type="n", xlab = paste0("X", i), ylab = paste0("X", j))
      text(X[[i]],X[[j]], cex = 0.7, pos=2)}
    else{
      hist(X[[i]], xlab = paste0("X", i), 
           main = paste0("Histogram of X", i), prob=TRUE,cex.main=0.8)
      lines(density(X[[i]]))}
  }
}

par(mfrow=c(1, 3), mar=c(2,2,2,2))
q <- vector("list", length = 3)
for (i in 1:3) {
  q[[i]]=qqnorm(X[[i]],type="n", main = paste0("Normal Q-Q Plot of X",i));qqline(X[[i]],col="red")
  text(q[[i]]$x, q[[i]]$y,  pos = 1)
}

#c
D <- c() 
for (i in 1:nrow(X)) {
  x_i <- t(X[i,])
  D[i]=t(x_i-x_bar)%*%solve(S)%*%(x_i-x_bar)
}
round(D,3)
plot(D, main = "D values vs index", type="n");text(D)

#e
new_X <- X[-c(5,12,54),]
S_new = cov(new_X)
S_new; round(S-S_new,4)
new_x_bar = colMeans(new_X)
new_x_bar; round(x_bar-new_x_bar,4)

#f
par(mfcol=c(3, 3), mar=c(2,2,2,2))
for(i in 1:3){
  for(j in 1:3) {
    if(i != j){
      plot(new_X[[i]],new_X[[j]],xlab = paste0("X", i), ylab = paste0("X", j))
    }
    else{
      hist(new_X[[i]], xlab = paste0("X", i), 
           main = paste0("Histogram of X", i), prob=TRUE,cex.main=0.8)
      lines(density(X[[i]]))}
  }
}
par(mfrow=c(1, 3), mar=c(2,2,2,2))
for (i in 1:3) {
  qqnorm(new_X[[i]], main = paste0("Normal Q-Q Plot of new X",i));qqline(X[[i]],col="red")
}
```




```{r, echo=FALSE, results='hide'}
CJ <- function (matrix) {
  s = ""
  for (i in 1:nrow(matrix)) {
    p <- paste(paste(as.character(matrix[i,]), collapse = " & "), "\\ ")
    s <- paste(s,p)
  }
  s <- paste("\begin{align} Y &= \begin{bmatrix}", s, "\\end{bmatrix} \\end{align}")
  # editing required afterwards
  return (s)
}
CJ(round(S-S_new,4))
```