---
title: "A5P1"
author: "Dawu Liu"
output:
  pdf_document: default
  word_document: default
---

```{r, echo=FALSE, results='hide', message=FALSE,warning = FALSE}
library(readxl)
library(factoextra)
X <- read_excel("C:/Users/John/Desktop/STAT 445/Data/temperaturedata-clean.xlsx")
```
In this assignment, principal component is written as \textbf{PC} sometimes for short.

Also, here are the keys for temperature data that will be analyzed below:

x1 = maximum daily air temperature

x2 = minimum daily air temperature

x3 = integrated area under daily air temperature curve

x4 = maximimum daily soil temperature

x5 = mininum daily soil temperature

x6 = integrated area under soil temperature

x7 = maximum daily relative humidity

x8 = minimum daily relative humidity

x9 = integrated area under daily humidity

x10 = total wind, miles per day

x11 = evaporation

\textbf{(a)} \

Sample covariance matrix \textbf{S}
```{r, echo = FALSE}
S <- cov(X)
round(S,4)
```

Sample correlation matrix \textbf{R}
```{r, echo = FALSE}
R <- cor(X)
round(R,4)
```
\textbf{(b) and (c)}\

PCA using covariance matrix \textbf{S} 
```{r, echo = FALSE}
S_pr = prcomp(X, center = TRUE, scale. = FALSE)
S_eigen_table = get_eigenvalue(S_pr)
```
\
i. The eigenvalues are:
```{r, echo = FALSE}
round(S_eigen_table[,1], 4)
```
The first eigenvalue is massively larger compared to the rest, and it accounts for the most of the total variance. 
(table shown in ii)

\
ii.

Criteria 1, eigenvalues and their cumulative proportions table
```{r, echo = FALSE}
round(S_eigen_table,4)
```
This method suggests to keep \textbf{1} principal component which gives 91.5% of the total variance.\

Criteria 2, check which eigenvalue(s) is greater than the mean of eigenvalues

The mean of the eigenvalues is:
```{r, echo = FALSE}
mean(S_eigen_table[,1])
```
This method also suggests to keep \textbf{1} principal component. \

Criteria3, scree plot

```{r,echo = FALSE,fig.height=4, fig.width=4}
plot(S_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
```

The "bend" occurs at PC2, indicating from PC2 and on, the the eigenvalues are relatively small. 
This method also suggests to keep \textbf{1} principal component. \

Overall, \textbf{1} principal component is retained.

But in order to make a scatter plot, we will use 2 principal components.\

\
iii.

The eigenvectors for the principal components:
```{r, echo = FALSE}
e_S1 <- as.vector(S_pr$rotation[,1])
e_S2 <- as.vector(S_pr$rotation[,2])
```
$(\tilde{\hat{e_1}})^T$ = 
```{r, echo = FALSE}
round(e_S1, 4)
```

$(\tilde{\hat{e_2}})^T$ = 
```{r, echo = FALSE}
round(e_S2, 4)
```
\
iv.

Almost all of PC1 depends on $x_{10}$, representing total wind (miles per day).
For PC2, we see $x_3$, $x_6$, $x_9$, and $x_{11}$ have significantly large magnitudes than the rest, 
means the majority of PC2 depends on integrated areas under the temperature/humidity curves and the evaporation. 
Also, PC2 shows the contrast between integrated area under daily humidity on one hand, and integrated 
area under daily air temperature, soil temperature, and evaporation on the other hand.

\
v.

Scatter plot for PC2 vs PC1

```{r, echo = FALSE, fig.height=5, fig.width=6}
plot(S_pr$x[,1], S_pr$x[,2], xlab = "PC1 (91.48%)", ylab = "PC2 (6.52%)", pch = 20)
text(S_pr$x[,1], S_pr$x[,2], pos = 3, offset = 0.3, cex = 0.5)
abline(v = 0, h = 0)
```

The density of the plot becomes higher as the PC1 value increases and PC2 value decreases. About half of the data points 
are grouped toward the right bottom portion of the plot.


\
PCA using sample correlation matrix \textbf{R}
```{r, echo = FALSE}
R_pr = prcomp(X, center = TRUE, scale. = TRUE)
R_eigen_table = get_eigenvalue(R_pr)
```
\
i. The eigenvalues are:
```{r, echo = FALSE}
round(R_eigen_table[,1], 4)
```
The first two eigenvalues are relatively larger compared to the rest, and they account for the majority of the total variance. 
(table shown in ii)

\
ii.

Criteria 1, eigenvalues and their cumulative proportions table
```{r, echo = FALSE}
round(R_eigen_table,4)
```
This method suggests to keep \textbf{3} principal component which gives 84.3% of the total variance.\

Criteria 2, check which eigenvalue(s) is greater than the mean of eigenvalues

Since three eigenvalues are larger than the mean of 1, this method suggests to keep \textbf{3} principal component. \

Criteria3, scree plot

```{r,echo = FALSE,fig.height=3, fig.width=4.2}
plot(R_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
```

The "bend" occurs at PC2, indicating from PC2 and on, the the eigenvalues are relatively small. 
This method suggests to keep \textbf{1} principal component. \

Overall, \textbf{3} principal components are retained. \

\
iii.

The eigenvectors for the principal components:
```{r, echo = FALSE}
e_R1 <- as.vector(R_pr$rotation[,1])
e_R2 <- as.vector(R_pr$rotation[,2])
e_R3 <- as.vector(R_pr$rotation[,3])
```
$(\tilde{\hat{e_1}})^T$ = 
```{r, echo = FALSE}
round(e_R1, 4)
```

$(\tilde{\hat{e_2}})^T$ = 
```{r, echo = FALSE}
round(e_R2, 4)
```

$(\tilde{\hat{e_3}})^T$ = 
```{r, echo = FALSE}
round(e_R3, 4)
```

\
iv.

PC1 depends on all the variables other than total wind(miles per day) and maximum daily relative humidity almost evenly.
Coefficients on humidity have opposite signs compared to temperature, showing contrast relationship.

The majority of PC2 depends on $x_5$, $x_8$, and $x_{10}$ have relatively large magnitudes than the rest,
means PC2 primarily interprets the minimum daily soil temperature, minimum daily relative humidity, and 
total wind(miles per day).

The majority of PC3 depends on maximum daily relative humidity, followed by a significant amount of 
total wind(miles per day) but with opposite sign, indicating there is contrast between those two.

\
v.

Scatter plot for the principal components

```{r, echo=FALSE, results='hide',fig.height=2.8, fig.width=8.4}
par(mfrow=c(1, 3))
plot(R_pr$x[,1], R_pr$x[,2], xlab = "PC1 (54.73%)", ylab = "PC2 (19.27%)", pch = 20)
abline(v = 0, h = 0)

plot(R_pr$x[,1], R_pr$x[,3], xlab = "PC1 (54.73%)", ylab = "PC3 (10.28%)", pch = 20)
abline(v = 0, h = 0)

plot(R_pr$x[,2], R_pr$x[,3], xlab = "PC2 (19.27%)", ylab = "PC3 (10.28%)", pch = 20)
abline(v = 0, h = 0)
```
PC2 VS PC1 plot: Data points are more gathered toward to the right side of the plot, data points becomes 
more scattered as they decrease in PC1 values.

PC3 VS PC1 plot: Data points are nearly evenly distributed across the plot, with some sparse points on the left, 
it's hard to get any relationships here.

PC2 VS PC3: It appears there are two small groups on the plots, one is slight above the intersection of the 
axes and one is slightly below the origin. 

\textbf{(d)} \

The boxplot of the data:

```{r, echo = FALSE,fig.height=5, fig.width=8} 
boxplot(X, ylab = 'measurement')
```
Correlation matrix \textbf{R} is better here. In the boxplot of the data above, the range for the values of $x_{10}$ are
significantly larger than the rest of the variables, causing a scaling issue.  In the \textbf{S} PCA, all
three criteria suggest to keep only one principal component, and that principal component only really 
tells us about the total wind(miles per day). Therefore, we can not obtain information about the relationships 
between the other variables from it. But in the \textbf{R} PCA, we are able to obtain those information.

\textbf{(e)} \ 

Interpretation of the principal components: 

For \textbf{S}:

PC1 is a component of total wind(miles per day), higher PC1 values means less wind(negative sign). \
PC2 shows the contrast between integrated area under daily humidity on one hand, 
and integrated area under daily air temperature/soil temperature, and evaporation on the other hand.

For \textbf{R}:

PC1 represents the contrast between humidity and temperature. \
PC2 mainly represents the minimum temperature/humidity values and the total wind (negative signs), 
higher PC2 values indicates lower minimum temperature/humidity, and less wind. Showing dependencies 
between those variables.\
PC3 shows the contrast between maximum daily relative humidity and total wind.  \

\
Code used to solve the questions(graphs are hidden):
```{r fig.show='hide'}
rm(list = ls())
library(readxl)
library(factoextra)
X <- read_excel("C:/Users/John/Desktop/STAT 445/Data/temperaturedata-clean.xlsx")
S <- cov(X)
round(S,4)
R <- cor(X)
round(R,4)

# S analysis
S_pr = prcomp(X, center = TRUE, scale. = FALSE)
S_eigen_table = get_eigenvalue(S_pr)
round(S_eigen_table[,1], 4)
round(S_eigen_table,4)
mean(S_eigen_table[,1])
plot(S_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
e_S1 <- as.vector(S_pr$rotation[,1])
e_S2 <- as.vector(S_pr$rotation[,2])
round(e_S1, 4)
round(e_S2, 4)
plot(S_pr$x[,1], S_pr$x[,2], xlab = "PC1 (91.48%)", ylab = "PC2 (6.52%)", pch = 20)
text(S_pr$x[,1], S_pr$x[,2], pos = 3, offset = 0.3, cex = 0.5)
abline(v = 0, h = 0)

# R analysis
R_pr = prcomp(X, center = TRUE, scale. = TRUE)
R_eigen_table = get_eigenvalue(R_pr)
round(R_eigen_table[,1], 4)
round(R_eigen_table,4)
plot(R_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
e_R1 <- as.vector(R_pr$rotation[,1])
e_R2 <- as.vector(R_pr$rotation[,2])
e_R3 <- as.vector(R_pr$rotation[,3])
round(e_R1, 4)
round(e_R2, 4)
round(e_R3, 4)
par(mfrow=c(1, 3))
plot(R_pr$x[,1], R_pr$x[,2], xlab = "PC1 (54.73%)", ylab = "PC2 (19.27%)", pch = 20)
abline(v = 0, h = 0)

plot(R_pr$x[,1], R_pr$x[,3], xlab = "PC1 (54.73%)", ylab = "PC3 (10.28%)", pch = 20)
abline(v = 0, h = 0)

plot(R_pr$x[,2], R_pr$x[,3], xlab = "PC2 (19.27%)", ylab = "PC3 (10.28%)", pch = 20)
abline(v = 0, h = 0)
```
