---
title: "A5P2"
output: pdf_document
author: Dawu Liu
---

```{r, echo=FALSE, results='hide', message=FALSE,warning = FALSE}
library(readxl)
library(factoextra)
data <- read_excel("C:/Users/John/Desktop/STAT 445/Data/fleabeetledata.xlsx")
data <- as.data.frame(data[-(1:2),-1])
```
In this assignment, principal component is written as \textbf{PC} sometimes for short.

Also, here are the keys for Flea Beetle data that will be analyzed below:

Distance TG = distance of transverse groove from posteriori border of prothorax

Elytra = length of elytra

Second Antenna = length of second antennal joint

Third Antenna = length of third antennal joint

\textbf{(a)} \
\textbf{PCA on Haltica oleracea group}

```{r, echo = FALSE}
X1 <- sapply(data[-20,1:4], as.numeric) 
colnames(X1) <- c("Distance TG", "Elytra", "Second Antenna", "Third Antenna")
X1_pr = prcomp(X1, center = TRUE, scale. = FALSE)
X1_eigen_table = get_eigenvalue(X1_pr)
```

\
i. Sample covariance matrix \textbf{S}
```{r, echo = FALSE}
S1 <- cov(X1)
round(S1,4)
```

\
ii. The eigenvalues are:
```{r, echo = FALSE}
round(X1_eigen_table[,1], 4)
```
The first two eigenvalues are relatively large compared to the rest. 
The first eigenvalue accounts for the majority of the total variance. (table shown in iii)

\
iii.

Criteria 1, eigenvalues and their cumulative proportions table
```{r, echo = FALSE}
round(X1_eigen_table,4)
```
This method suggests to keep \textbf{2} principal component which gives 87.0% of the total variance.\

Criteria 2, check which eigenvalue(s) is greater than the mean of eigenvalues

The mean of the eigenvalues is:
```{r, echo = FALSE}
mean(X1_eigen_table[,1])
```
This method suggests to keep \textbf{1} principal component. \

Criteria3, scree plot

```{r,echo = FALSE,fig.height=4, fig.width=4}
plot(X1_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
```

The "bend" occurs at PC2 is the strongest(the bend at PC3 is only slight weaker), indicating from PC2 and on, the the eigenvalues are relatively small. 
This method suggests to keep \textbf{1} principal component. 

Overall, the criteria suggest \textbf{1} principal component should be retained. This is a very close call, 
since criteria 1 suggests to keep 2, criteria 2 suggests to keep 1, and in criteria 3 the angles at PC2 and PC3 are nearly identical.

But in order to make a scatter plot, we will use 2 principal components and proceed.\

\
iv. The eigenvectors for the principal components:

```{r, echo = FALSE}
e_X1_1 <- X1_pr$rotation[,1]
e_X1_2 <- X1_pr$rotation[,2]
```
$(\tilde{\hat{e_1}})^T$ = 
```{r, echo = FALSE}
round(e_X1_1, 4)
```

$(\tilde{\hat{e_2}})^T$ = 
```{r, echo = FALSE}
round(e_X1_2, 4)
```
\
v. 

PC1 depends on the length of elytra the most, followed by the distance of transverse groove from 
posteriori border of prothorax and the length of third antennal joint. All four variables contribute 
a considerable amount to PC1, indicating PC1 is related to the size of the beetles.

PC2 depends on the length of third antennal joint the most, then the length of elytra but in opposite 
direction, showing there is some contrast between those two. This indicats PC2 is related to the shape of the beetles. 

\
vi. Scatter plot for PC2 vs PC1

```{r,echo = FALSE,fig.height=4, fig.width=4}
plot(X1_pr$x[,1], X1_pr$x[,2], xlab = "PC1 (66.88%)", ylab = "PC2 (22.13%)", pch = 20)
text(X1_pr$x[,1], X1_pr$x[,2], pos = 3, offset = 0.3, cex = 0.5)
abline(v = 0, h = 0)
```

The data appear to have a weak linear relationship. The data can somewhat fit into an ellipse shape. 
Observation number 4 on the far right for example, has the largest PC1 value, indicating it is the largest beetle in size. \

\textbf{PCA on Haltica carduorum group}

```{r, echo = FALSE}
X2 <- sapply(data[,5:8], as.numeric) 
colnames(X2) <- c("Distance TG", "Elytra", "Second Antenna", "Third Antenna")
X2_pr = prcomp(X2, center = TRUE, scale. = FALSE)
X2_eigen_table = get_eigenvalue(X2_pr)
```

\
i. Sample covariance matrix \textbf{S}
```{r, echo = FALSE}
S2 <- cov(X2)
round(S2,4)
```

\
ii. The eigenvalues are:
```{r, echo = FALSE}
round(X2_eigen_table[,1], 4)
```
The first two eigenvalues are relatively large compared to the rest. 
The first eigenvalue accounts for the majority of the total variance. (table shown in iii)

\
iii.

Criteria 1, eigenvalues and their cumulative proportions table
```{r, echo = FALSE}
round(X2_eigen_table,4)
```
This method suggests to keep \textbf{2} principal component which gives 83.8% of the total variance.\

Criteria 2, check which eigenvalue(s) is greater than the mean of eigenvalues

The mean of the eigenvalues is:
```{r, echo = FALSE}
mean(X2_eigen_table[,1])
```
This method suggests to keep \textbf{1} principal component. \

Criteria3, scree plot

```{r,echo = FALSE,fig.height=3.5, fig.width=4}
plot(X2_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
```

The "bend" occurs at PC2, indicating from PC2 and on, the the eigenvalues are relatively small. 
This method suggests to keep \textbf{1} principal component. \

Overall, the criteria suggest \textbf{1} principal component should be retained.

But in order to make a scatter plot, we will use 2 principal components.\
\
iv. The eigenvectors for the principal components:

```{r, echo = FALSE}
e_X2_1 <- X2_pr$rotation[,1]
e_X2_2 <- X2_pr$rotation[,2]
```
$(\tilde{\hat{e_1}})^T$ = 
```{r, echo = FALSE}
round(e_X2_1, 4)
```

$(\tilde{\hat{e_2}})^T$ = 
```{r, echo = FALSE}
round(e_X2_2, 4)
```
\
v. 

PC1 depends on the length of elytra the most, then depends on the rest of the three variables almost evenly. 
All four variables contribute a considerable amount to PC1 in the same direction, indicating PC1 is 
related to the size of the beetles. 

PC2 primarily represents the length of third antennal joint. Distance TG and the length of elystra have a opposite sign to it, show there are some contrast between those three.

\
vi. Scatter plot for PC2 vs PC1

```{r,echo = FALSE,fig.height=4, fig.width=4}
plot(X2_pr$x[,1], X2_pr$x[,2], xlab = "PC1 (66.45%)", ylab = "PC2 (17.39%)", pch = 20, col = 2)
text(X2_pr$x[,1], X2_pr$x[,2], pos = 3, offset = 0.3, cex = 0.5)
abline(v = 0, h = 0)
```

Observation number 2 on the far left has the smallest PC1 values, indicating it is the smallest beetle in size, 
appears to be an outlier. Observation number 18 on the top for example, has long third antenna, but probably short elytra. There is a gap around the origin which in the the middle between the observations, which might due to small sample size. \

\textbf{PCA using data from both groups}
```{r, echo = FALSE}
X <-rbind(X1,X2)
X_pr = prcomp(X, center = TRUE, scale. = FALSE)
X_eigen_table = get_eigenvalue(X_pr)
```
\
i. Sample covariance matrix \textbf{S}
```{r, echo = FALSE}
S <- cov(X)
round(S,4)
```

\
ii. The eigenvalues are:
```{r, echo = FALSE}
round(X_eigen_table[,1], 4)
```
The first eigenvalue is relatively large and contributes to the majority of the total variance. (table shown in iii)

\
iii.

Criteria 1, eigenvalues and their cumulative proportions table
```{r, echo = FALSE}
round(X_eigen_table,4)
```
This method suggests to keep \textbf{2} principal component which gives 84.02% of the total variance.\

Criteria 2, check which eigenvalue(s) is greater than the mean of eigenvalues

The mean of the eigenvalues is:
```{r, echo = FALSE}
mean(X_eigen_table[,1])
```
This method suggests to keep \textbf{1} principal component. \

Criteria3, scree plot

```{r,echo = FALSE,fig.height=4, fig.width=4}
plot(X_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
```

The "bend" occurs at PC2, indicating from PC2 and on, the the eigenvalues are relatively small. 
This method suggests to keep \textbf{1} principal component. \

Overall, the criteria suggest \textbf{1} principal component should be retained.

But in order to make a scatter plot, we will use 2 principal components.\

\
iv. The eigenvectors for the principal components:

```{r, echo = FALSE}
e_X_1 <- X_pr$rotation[,1]
e_X_2 <- X_pr$rotation[,2]
```
$(\tilde{\hat{e_1}})^T$ = 
```{r, echo = FALSE}
round(e_X_1, 4)
```

$(\tilde{\hat{e_2}})^T$ = 
```{r, echo = FALSE}
round(e_X_2, 4)
```
\
v. 

PC1 strongly depends on the length of elytra, followed by the length of third and second antenna. There is 
almost no dependency on Distance TG. Since all the coefficients are in the same directions, 
PC1 might be an indication of the size of the beetles.

PC2 primarily depends on distance TG. It also shows opposite relationship between 
distance TG, the length of elytra on one hand, and the length of second, third antenna on another hand.

\
vi. Scatter plot for PC2 vs PC1

```{r,echo = FALSE,fig.height=4, fig.width=4}
PR <- cbind(X_pr$x[,1], X_pr$x[,2])
PR <- cbind(PR,c(1))
PR[20:39,3] <- 2
plot(PR[,1], PR[,2], xlab = "PC1 (65.07%)", ylab = "PC2 (18.95%)", pch = 20, col = PR[,3])
text(PR[,1], PR[,2], pos = 4, offset = 0.2, cex = 0.5)
abline(v = 0, h = 0)
par(xpd = TRUE)
legend("topright", legend=c("Haltica oleracea", "Haltica carduorum"), 
       lty= c(0,0), pch=c(20, 20),col = c(1,2), cex= 0.5)
```

Haltica oleracea have higher average values in both PC1 and PC2 than Haltica carduorum, it appear those two group belongs to 
two different clusters. And each group data appears to somewhat fit in an ellipse. Since the values of the 
eigenvectors for PC1 are all negative, this shows Haltica oleracea tends to be slightly smaller than Haltica carduorum in size. 
Both groups have a negative relationship between PC1 and PC2. \

\textbf{(b)}

Each type of the analysis has their own advantages. Since all four variables are the measurements of the length 
on certain body parts, the first two analyses give more accurate information on the relationships between body sizes within each group. 
The third analysis (using combined data) shows how one group is related to another, 
telling us there is a clear divide between the two groups. \

\
Code used to solve the questions(graphs are hidden):

```{r fig.show='hide'}
rm(list = ls())
library(readxl)
library(factoextra)
data <- read_excel("C:/Users/John/Desktop/STAT 445/Data/fleabeetledata.xlsx")
data <- as.data.frame(data[-(1:2),-1])

# group 1
X1 <- sapply(data[-20,1:4], as.numeric) 
colnames(X1) <- c("Distance TG", "Elytra", "Second Antenna", "Third Antenna")
X1_pr = prcomp(X1, center = TRUE, scale. = FALSE)
X1_eigen_table = get_eigenvalue(X1_pr)
S1 <- cov(X1)
round(S1,4)
round(X1_eigen_table[,1], 4)
round(X1_eigen_table,4)
mean(X1_eigen_table[,1])
plot(X1_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
e_X1_1 <- X1_pr$rotation[,1]
e_X1_2 <- X1_pr$rotation[,2]
round(e_X1_1, 4)
round(e_X1_2, 4)
plot(X1_pr$x[,1], X1_pr$x[,2], xlab = "PC1 (66.88%)", ylab = "PC2 (22.13%)", pch = 20)
text(X1_pr$x[,1], X1_pr$x[,2], pos = 3, offset = 0.3, cex = 0.5)
abline(v = 0, h = 0)

# group2
X2 <- sapply(data[,5:8], as.numeric) 
colnames(X2) <- c("Distance TG", "Elytra", "Second Antenna", "Third Antenna")
X2_pr = prcomp(X2, center = TRUE, scale. = FALSE)
X2_eigen_table = get_eigenvalue(X2_pr)
S2 <- cov(X2)
round(S2,4)
round(X2_eigen_table[,1], 4)
round(X2_eigen_table,4)
mean(X2_eigen_table[,1])
plot(X2_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
e_X2_1 <- X2_pr$rotation[,1]
e_X2_2 <- X2_pr$rotation[,2]
round(e_X2_1, 4)
round(e_X2_2, 4)
plot(X2_pr$x[,1], X2_pr$x[,2], xlab = "PC1 (66.45%)", ylab = "PC2 (17.39%)", pch = 20, col = 2)
text(X2_pr$x[,1], X2_pr$x[,2], pos = 3, offset = 0.3, cex = 0.5)
abline(v = 0, h = 0)

# combined
X <-rbind(X1,X2)
X_pr = prcomp(X, center = TRUE, scale. = FALSE)
X_eigen_table = get_eigenvalue(X_pr)
S <- cov(X)
round(S,4)
round(X_eigen_table[,1], 4)
round(X_eigen_table,4)
mean(X_eigen_table[,1])
plot(X_eigen_table[,1], type = "o", pch = 15, main = "Scree Plot",
     xlab = "Principal Component Number", ylab = "Eigenvalues")
e_X_1 <- X_pr$rotation[,1]
e_X_2 <- X_pr$rotation[,2]
round(e_X_1, 4)
round(e_X_2, 4)
PR <- cbind(X_pr$x[,1], X_pr$x[,2])
PR <- cbind(PR,c(1))
PR[20:39,3] <- 2
plot(PR[,1], PR[,2], xlab = "PC1 (65.07%)", ylab = "PC2 (18.95%)", pch = 20, col = PR[,3])
text(PR[,1], PR[,2], pos = 4, offset = 0.2, cex = 0.5)
abline(v = 0, h = 0)
par(xpd = TRUE)
legend("topright", legend=c("Haltica oleracea", "Haltica carduorum"), 
       lty= c(0,0), pch=c(20, 20),col = c(1,2), cex= 0.5)
```