---
title: "Assignment06"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(readxl)
library(psych)
protein <- read_excel("C:/Users/John/Desktop/STAT 445/Data/europe_protein_data.xls")[,-1]
R <- cor(protein)
maketable <- function(){
  table <- matrix(NA, nrow=11,ncol=6)
  rownames(table) <- c(colnames(R), "Var. Acc. For", "Prop. Tot. Var.")
  colnames(table) <- c("load 1", "load 2", "load 3", "load 4", "commun.", "spec.var.")
  return(table)
}

filltable <- function(t, x) {
  for(i in 1:4){
    t[1:9,i] <- x$loadings[,i]
    t[10:11,i] <- x$Vaccounted[1:2,i]
  }
  t[1:9,5] <- x$communality
  t[1:9,6] <- x$uniquenesses
  return(t)
}

```

4 factors and correlation matrix R are used for all four analyses. 
We assume the data is normal in order to make maximum likelihood method work. 
There are 25 observations for each of the 9 variables. 
Results have been round to 4 decimal places.




\textbf{(a) Principal components method without rotation.}

\
i. Display the table of results
```{r echo=FALSE}
a <- principal(R,nfactors=4, n.obs=25, rotate="none")
t1 <- maketable()
t1 <- filltable(t1,a)
round(t1, 4)
```
\
ii. Show the estimate $\hat{L}\hat{L}^T+\hat{\Psi}$
```{r echo=FALSE}
estimate1 <-a$loadings%*%t(a$loadings)+diag(a$uniquenesses)
colnames(estimate1)=NULL; rownames(estimate1)=NULL
round(estimate1,4)
```
\
iii. Show the error matrix $E=R-(\hat{L}\hat{L}^T+\hat{\Psi})$ and $\|{E}\|$

E = 
```{r echo=FALSE}
error1 <- R-estimate1
colnames(error1)=NULL; rownames(error1)=NULL
round(error1,4)
```

$\|{E}\|$ = 
```{r echo=FALSE}
round(sqrt(sum(error1^2)),4)
```




\textbf{(b) Principal components method with varimax rotation.}

\
i. Display the table of results
```{r echo=FALSE}
b <- principal(R,nfactors=4, n.obs=25, rotate="varimax")
t2 <- maketable()
t2 <- filltable(t2,b)
round(t2, 4)
```
\
ii. Show the estimate $\hat{L}\hat{L}^T+\hat{\Psi}$
```{r echo=FALSE}
estimate2 <-b$loadings%*%t(b$loadings)+diag(b$uniquenesses)
colnames(estimate2)=NULL; rownames(estimate2)=NULL
round(estimate2,4)
```
\
iii. Show the error matrix $E=R-(\hat{L}\hat{L}^T+\hat{\Psi})$ and $\|{E}\|$

E = 
```{r echo=FALSE}
error2 <- R-estimate2
colnames(error2)=NULL; rownames(error2)=NULL
round(error2,4)
```

$\|{E}\|$ = 
```{r echo=FALSE}
round(sqrt(sum(error2^2)),4)
```





\textbf{(c) Maximum likelihood method without varimax rotation.}
\
i. Display the table of results
```{r echo=FALSE}
c <- fa(R,nfactors=4, n.obs=25, rotate="none", fm="ml")
t3 <- maketable()
t3 <- filltable(t3,c)
round(t3, 4)
```
\
ii. Show the estimate $\hat{L}\hat{L}^T+\hat{\Psi}$
```{r echo=FALSE}
estimate3 <-c$loadings%*%t(c$loadings)+diag(c$uniquenesses)
colnames(estimate3)=NULL; rownames(estimate3)=NULL
round(estimate3,4)
```
\
iii. Show the error matrix $E=R-(\hat{L}\hat{L}^T+\hat{\Psi})$ and $\|{E}\|$

E = 
```{r echo=FALSE}
error3 <- R-estimate3
colnames(error3)=NULL; rownames(error3)=NULL
round(error3,4)
```

$\|{E}\|$ = 
```{r echo=FALSE}
round(sqrt(sum(error3^2)),4)
```




\textbf{(d) Maximum likelihood method with varimax rotation.}
\
i. Display the table of results
```{r echo=FALSE}
d <- fa(R,nfactors=4, n.obs=25, rotate="varimax", fm="ml")
t4 <- maketable()
t4 <- filltable(t4,d)
round(t4, 4)
```
\
ii. Show the estimate $\hat{L}\hat{L}^T+\hat{\Psi}$
```{r echo=FALSE}
estimate4 <-d$loadings%*%t(d$loadings)+diag(d$uniquenesses)
colnames(estimate4)=NULL; rownames(estimate4)=NULL
round(estimate4,4)
```
\
iii. Show the error matrix $E=R-(\hat{L}\hat{L}^T+\hat{\Psi})$ and $\|{E}\|$

E = 
```{r echo=FALSE}
error4 <- R-estimate4
colnames(error4)=NULL; rownames(error4)=NULL
round(error4,4)
```

$\|{E}\|$ = 
```{r echo=FALSE}
round(sqrt(sum(error4^2)),4)
```

\textbf{(e)}

\
i. \textbf{Both} methods produce a reasonable good approximation. Since rotation doesn't affect the error, 
there are only two error matrices, one for each method. 

Three criteria are used to check the errors:

1. Check $\|E\|$ and compare to $\|R\|$ ,where $\|R\|$ = 4.618.

2. Check the entries in error matrix and their relative error proportion, which is calculated by dividing 
each entry in the error matrix by its corresponding entry in the R matrix, then take absolute value of it.

3. Check the average (mean) of the relative error proportion.

```{r echo=FALSE, results = 'hide'}
length(which(abs(error1/R)<0.15))
length(which(abs(error3/R)<0.15))
round(sqrt(sum(R^2)),4)
mean(abs(error1/R))
mean(abs(error3/R))
```

For principal components method error matrix: 

1. $\|E\|$ is relatively small compared to $\|R\|$. \
2. \textbf{53} out of \textbf{81} entries has an error \textbf{below 15\%}. \
3. Average relative error is \textbf{17.12\%}, still reasonable. 

For maximum likelihood method error matrix:

1. $\|E\|$ is relatively small compared to $\|R\|$. \
2. \textbf{71} out of \textbf{81} entries has an error \textbf{below 15\%} \
3. Average relative error is \textbf{8.07\%}, which is pretty reasonable.

Therefore, \textbf{both} methods produce a reasonable good approximation.  \

\
ii. \textbf{Maximum likelihood method} produces the better approximation here. It has the smaller $\|E\|\approx 0.22$ and 
average relative error %. 
Most of the entries in the maximum likelihood error matrix have \textbf{smaller maginitudes} 
than their counter part in the principal components error matrix. We can observe this by 
\textbf{subtracting the absolute values} of the matrices: \
abs(E(principal components)) - abs(E(maximum likelihood)) = 

```{r echo=FALSE}
round(abs(error1)-abs(error3),4)
```

Nearly all the entries of the result are positive, meaning the entries in maximum likelihood 
error matrix have \textbf{smaller errors} in general.


\newpage
Code used to solve the questions:
```{r message=FALSE, warning=FALSE}
rm(list = ls())
library(readxl)
library(psych)
protein <- read_excel("C:/Users/John/Desktop/STAT 445/Data/europe_protein_data.xls")[,-1]
R <- cor(protein)
maketable <- function(){
  table <- matrix(NA, nrow=11,ncol=6)
  rownames(table) <- c(colnames(R), "Var. Acc. For", "Prop. Tot. Var.")
  colnames(table) <- c("load 1", "load 2", "load 3", "load 4", "commun.", "spec.var.")
  return(table)
}

filltable <- function(t, x) {
  for(i in 1:4){
    t[1:9,i] <- x$loadings[,i]
    t[10:11,i] <- x$Vaccounted[1:2,i]
  }
  t[1:9,5] <- x$communality
  t[1:9,6] <- x$uniquenesses
  return(t)
}

# a
# i 
a <- principal(R,nfactors=4, n.obs=25, rotate="none")
t1 <- maketable()
t1 <- filltable(t1,a)
round(t1, 4)
# ii
estimate1 <-a$loadings%*%t(a$loadings)+diag(a$uniquenesses)
colnames(estimate1)=NULL; rownames(estimate1)=NULL
round(estimate1,4)
# iii
error1 <- R-estimate1
colnames(error1)=NULL; rownames(error1)=NULL
round(error1,4)
round(sqrt(sum(error1^2)),4)

# b
# i
b <- principal(R,nfactors=4, n.obs=25, rotate="varimax")
t2 <- maketable()
t2 <- filltable(t2,b)
round(t2, 4)
# ii
estimate2 <-b$loadings%*%t(b$loadings)+diag(b$uniquenesses)
colnames(estimate2)=NULL; rownames(estimate2)=NULL
round(estimate2,4)
# iii
error2 <- R-estimate2
colnames(error2)=NULL; rownames(error2)=NULL
round(error2,4)
round(sqrt(sum(error2^2)),4)

# c
# i
c <- fa(R,nfactors=4, n.obs=25, rotate="none", fm="ml")
t3 <- maketable()
t3 <- filltable(t3,c)
round(t3, 4)
# ii
estimate3 <-c$loadings%*%t(c$loadings)+diag(c$uniquenesses)
colnames(estimate3)=NULL; rownames(estimate3)=NULL
round(estimate3,4)
# iii
error3 <- R-estimate3
colnames(error3)=NULL; rownames(error3)=NULL
round(error3,4)
round(sqrt(sum(error3^2)),4)

# d
# i
d <- fa(R,nfactors=4, n.obs=25, rotate="varimax", fm="ml")
t4 <- maketable()
t4 <- filltable(t4,d)
round(t4, 4)
# ii
estimate4 <-d$loadings%*%t(d$loadings)+diag(d$uniquenesses)
colnames(estimate4)=NULL; rownames(estimate4)=NULL
round(estimate4,4)
# iii
error4 <- R-estimate4
colnames(error4)=NULL; rownames(error4)=NULL
round(error4,4)
round(sqrt(sum(error4^2)),4)

# e
# i
length(which(abs(error1/R)<0.15))
length(which(abs(error3/R)<0.15))
round(sqrt(sum(R^2)),4)
mean(abs(error1/R))
mean(abs(error3/R))
# ii
round(abs(error1)-abs(error3),4)



